Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-29 bbbq_article.tex
Fix spelling errors

This study has two objectives. First, we aimed to generalize our findings by predicting the presentation of peptides from different kingdoms of life and for both MHC-I and -II. From these in silico predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, and for most haplotypes of both MHC-I and II. We confirmed this over-presentation of TMH-derived peptides by re-analyzis of a peptide elution study. Second, we tested our hypothesis that TMHs
are more evolutionarily conserved than  solvent-exposed protein regions. Our analysis of human
single nucleotide polymorphisms (SNPs) showed that random point mutations are indeed less likely
to occur within TMHs. These findings strengthen the emerging notion that TMHs are important for the T cell branch of the adaptive immune system, and hence are of  
overlooked importance in vaccine development.

@@

The number of SNPs
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analysed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

@@

Figure \ref{fig:1} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
for at least one residue.
A dashed line indicates the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as soluble regions.
In 12 out of 13 MHC-I haplotypes, TMH-derived epitopes are presented more often 
then the null expectation, for a human proteome. 
For the viral and bacterial proteome, 11 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the various different haplotypes 
is similar for the probed proteomes, which strenghtens our previous conclusion (\cite{bianchi2017}) that the hydrophopbicity of the MHC-binding groove is the main factor responsible for the predicted over-presentation of TMH-derived peptides.

@@

We next wondered if the over-representation of TMH-derived peptided would also be present for MHC-II. 
Figure \ref{fig:2} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, and this over-presentation seems even more pronounced than for MHC-I. 
\richel{
  Showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{TMH-derived epitopes are predicted to be presentated in vivo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevant zijn de haplotypen die veel TMH presenteren?}
The over-presentation of TMHs for most MHC-II haplotypes 
is more pronounced than for MHC-I haplotypes,
as up to a third of the predicted MHC-II binders 
\richel{Hier ben ik nu met verwerken...} 
from the human and MtB proteome stem from TMHs. \frans{ik zoek nog naar associaties} 
such as \verb;HLA-DQA1*0101/DQB1*0501;

@@

Evolutionary selection of pathogens by a host's immune system,
however, is likelier to occur for proteomic patterns that are general,
over patterns that are rare.
Where essential catalytic sites in a pathogenic proteome
are rare, TMHs are common and thus a more feasible 
target for evolution to respond to.
We have found the signature of evolution when both factors,
that is, TMHs and catalytic sites are likely to co-occur,
which is in proteins that span the membrane at least twice.
This may be explained additionally that multi-spoanners
are not only likelier to have a catalytically active site,
also there are interactions between the TMHs.
For single-spanners, we found no evolutionary conservation.
We speculate that the human immune system is more attentive 
towards TMHs in multi-spanners, as these are evolutionarily conserved.

>>>>

This study has two objectives. First, we aimed to generalize our findings by predicting the presentation of peptides from different kingdoms of life and for both MHC-I and -II. From these in silico predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, and for most haplotypes of both MHC-I and II. We confirmed this over-presentation of TMH-derived peptides by re-analysis of a peptide elution study. Second, we tested our hypothesis that TMHs
are more evolutionarily conserved than  solvent-exposed protein regions. Our analysis of human
single nucleotide polymorphisms (SNPs) showed that random point mutations are indeed less likely
to occur within TMHs. These findings strengthen the emerging notion that TMHs are important for the T cell branch of the adaptive immune system, and hence are of  
overlooked importance in vaccine development.

@@

The number of SNPs
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positions of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

@@

Figure \ref{fig:1} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
for at least one residue.
A dashed line indicates the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as soluble regions.
In 12 out of 13 MHC-I haplotypes, TMH-derived epitopes are presented more often 
then the null expectation, for a human proteome. 
For the viral and bacterial proteome, 11 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the various different haplotypes 
is similar for the probed proteomes, which strengthens our previous conclusion (\cite{bianchi2017}) that the hydrophobicity of the MHC-binding groove is the main factor responsible for the predicted over-presentation of TMH-derived peptides.

@@

We next wondered if the over-representation of TMH-derived peptides would also be present for MHC-II. 
Figure \ref{fig:2} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, and this over-presentation seems even more pronounced than for MHC-I. 
\richel{
  Showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{TMH-derived epitopes are predicted to be presented in vivo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevant zijn de haplotypen die veel TMH presenteren?}
The over-presentation of TMHs for most MHC-II haplotypes 
is more pronounced than for MHC-I haplotypes,
as up to a third of the predicted MHC-II binders 
\richel{Hier ben ik nu met verwerken...} 
from the human and MTb proteome stem from TMHs. \frans{ik zoek nog naar associaties} 
such as \verb;HLA-DQA1*0101/DQB1*0501;

@@

Evolutionary selection of pathogens by a host's immune system,
however, is likelier to occur for proteomic patterns that are general,
over patterns that are rare.
Where essential catalytic sites in a pathogenic proteome
are rare, TMHs are common and thus a more feasible 
target for evolution to respond to.
We have found the signature of evolution when both factors,
that is, TMHs and catalytic sites are likely to co-occur,
which is in proteins that span the membrane at least twice.
This may be explained additionally that multi-spanners
are not only likelier to have a catalytic active site,
also there are interactions between the TMHs.
For single-spanners, we found no evolutionary conservation.
We speculate that the human immune system is more attentive 
towards TMHs in multi-spanners, as these are evolutionarily conserved.
